---
title: "MCMC Demo 2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Exercise 
Consider the mixture distribution described on p. 99 (Hoff). This distribution is a joint probability distribution of a discrete variable $\delta = \{1,2,3\},$ denoting which mixture component the mass comes from and a continuous variable $\theta$. The target density is $\{Pr(\delta = 1), Pr(\delta=2), Pr(\delta=3)\} = (.45,.10,.45)$ and $p(\theta|\delta=i) \sim N(\theta; \mu_i,\sigma^2_i)$ where $\{\mu_1,\mu_2,\mu_3\} = (-3,0,3)$ and $\sigma^2_i = 1/3$ for $i \in \{1,2,3\}$.
\vfill

1. Generate 1000 samples of $\theta$ from this distribution using a Monte Carlo procedure. (Hint: first generate $\delta^{(i)}$ from the marginal distribution $p(\delta)$ and then generate $\theta^{(i)}$ from $p(\theta|\delta)$.) Plot your samples in a histogram form and superimpose a curve of the density function. Comment on your samples, do they closely match the true distribution?
\vfill

```{r}
num_sims <- 1000
delta <- theta <- rep(0, num_sims)
mu_vals <- c(-3,0,3)
sigma_vals <- rep(sqrt(1/3),3)

for (iter in 1:num_sims){
  delta[iter] <- sample(1:3, size = 1, prob = c(.45,.1,.45))
  
  # Now sample theta | delta 
}
```


2. Next, generate samples from a Gibbs sampler using the full conditional distributions of $\theta$ and $\delta$. You already know the form of the full conditional for $\theta$ from above. The full conditional distribution for $\delta$ is given below:
	$$Pr(\delta=d|\theta) = \frac{Pr(\delta=d) \times p(\theta|\delta=d)}{\sum_{d=1}^3Pr(\delta=d) \times p(\theta|\delta=d)}$$

	Hint: for $p(\theta|\delta=d)$ evaluate $\theta$ from a normal distribution with parameters $\{\mu_d,\sigma^2_d\}$. Intialize $\theta$ at 0.
\vfill	


\vfill
    
a. Generate 100 samples using this procedure. Plot your samples as a histogram with the true density superimposed on the plot. Also include a plot of your $\theta$ value on the y-axis and the iteration number on the x-axis. This is called a trace plot, and allows your to visualize the movement of your MCMC *particle*. Comment on how close your samples match the true density. What does the trace plot reveal about the position of $\theta$ over time (the iterations)? Does the proportion of the time the sample spends in each state ($\delta$) match the true probabilities?
    
```{r}
num_sims <- 100
delta <- rep(2, num_sims)
theta <- rep(0, num_sims)

for (iter in 2:num_sims){
  delta_probs <- c(.45 *  dnorm(theta[iter-1], mean = mu_vals[1], sd  = sigma_vals[1]),
                   .10 *  dnorm(theta[iter-1], mean = mu_vals[2], sd  = sigma_vals[2]),
                   .45 *  dnorm(theta[iter-1], mean = mu_vals[3], sd  = sigma_vals[3]))
  delta_probs <- delta_probs / sum(delta_probs)
  delta[iter] <- sample(1:3, size = 1, prob = delta_probs)
  
  # Now sample theta | delta 
}
```	    
b. Repeat for 1000 samples.
c. Repeat for 10000 samples.
